---
title: 'BDSI R Markdown: Exercise 3'
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

```{r}
library(tidyverse)
```


## Instructions

In these exercises you will explore the xenograft data. Create code chunks to do 
the necessary calculations and save your results. Write English prose to answer 
the questions.

## Read in the data

Make a new chunk below this sentence and inside it create an object called `tumor_growth` object by reading in the data using `read_csv`.

```{r}
tumor_growth <- read_csv("tumor_growth.csv")
```

Look over the code chunk below and confer with your group about what you think 
the code is doing. Feel free to look up any unfamiliar functions. Then run the 
chunk (make sure you've run the chunk that you wrote above first). 

## Calculate relative size

```{r}
tumor_growth <- 
  tumor_growth %>% 
  group_by(ID) %>% 
  arrange(Day) %>% 
  mutate(Size_Day0 = first(Size),
         relative_Size = log(Size / Size_Day0)) %>% 
  ungroup() %>% 
  arrange(ID)
```

## Calculate summary statistics

Separately for each treatment group, create code chunks to calculate for each 
of days 0, 7, and 14 the following statistics: the mean tumor size, the mean 
*relative* tumor size as defined in the chunk above, and the number of 
measurements. 

```{r}
tumor_growth_summarized <- 
  tumor_growth %>%
  filter(Day %in% c(0, 7, 14)) %>%
  group_by(Grp, Day) %>%
  summarize(mean_size = mean(Size), 
            mean_relative_size = mean(relative_Size), 
            number_measurements = n())
```

Report these statistics. Which group had the largest typical 
growth between day 0 and days 7 / 14? Which group had the most measurements 
taken on day 14? Why do you think the mice have different numbers of measurements 
taken on these days?

*The control group had the largest typical growth, with average relative growth values of 1.69 and 3.06 at days 7 and 14, respectively*

*The the drug-alone and radiation-alone treatment groups had the most measurements at day 14 (nine measurements) This is probably due to mice in the control group having been sacrificed earlier.*

**You could also use inline code chunks here. For example, I could write the first sentence above as:**

```{r, include = F}
# I use the glue_collapse function
library(glue);
```


*The control group had the largest typical growth, with average relative growth values of `r tumor_growth_summarized %>% filter(Day %in% c(7,14), Grp == "1.CTR") %>% pull(mean_relative_size) %>% round(2) %>% glue_collapse(sep = " and ")` at days 7 and 14, respectively*


*Hint* You might find the `n()` function to be useful. It counts the current group 
size (but it only works inside of `summarize` or `mutate`). Check the help page
if needed.

## Plots

Next, use `ggplot` to create a 'spaghetti plot' of the tumor growth data that
plots tumor size against day of the measurement. Each mouse should have its own
line (the spaghetti strand), and the strand should be colored by the treatment
group it is in. You will want to use `geom_line` and use `group = ID` as an
aesthetic to tell the geom that the values from the same ID should be connected
by the line. 

Create two spaghetti plots: one with `Size` variable on the $y$-axis and one with `relative_Size` variable on the $y$-axis. 


```{r}
ggplot(tumor_growth, 
       mapping = aes(x = Day, y = Size, color = Grp)) + 
  geom_line(aes(group = ID),
            # set alpha to < 1 so that individual mice are down-emphasized
            alpha = 0.25) + 
  # plot summary statistics with a thick line 
  # the aesthetics will be inherited from the ggplot
  geom_line(data = tumor_growth %>% group_by(Grp, Day) %>% summarize(Size = mean(Size)), 
            size = 1.25) + 
  labs(x = "Day of measurement", y = "Tumor size") + 
  # limit the viewing to the first 18 days, where we have more measurements per group
  coord_cartesian(xlim = c(0, 18)) + 
  theme(text = element_text(size = 16)) 
```


```{r}
ggplot(tumor_growth, 
       mapping = aes(x = Day, y = relative_Size, color = Grp)) + 
  geom_line(aes(group = ID),
            # set alpha to < 1 so that individual mice are down-emphasized
            alpha = 0.25) + 
  # plot summary statistics with a thick line 
  # the aesthetics will be inherited from the ggplot
  geom_line(data = tumor_growth %>% group_by(Grp, Day) %>% summarize(relative_Size = mean(relative_Size)), 
            size = 1.25) + 
  labs(x = "Day of measurement", y = "Relative tumor size") + 
  # limit the viewing to the first 18 days, where we have more measurements per group
  coord_cartesian(xlim = c(0, 18)) + 
  theme(text = element_text(size = 16)) 
```

Using some more advanced programming, I can also make a function that will
generate either plot:

```{r}
tumor_spaghetti <- 
  function(size_var, y_axis_name) {
    ggplot(tumor_growth, 
           mapping = aes(x = Day, y = !!sym(size_var), color = Grp)) + 
      geom_line(aes(group = ID),
                # set alpha to < 1 so that individual mice are down-emphasized
                alpha = 0.25) + 
      # plot summary statistics with a thick line 
      # the aesthetics will be inherited from the ggplot
      geom_line(data = tumor_growth %>% group_by(Grp, Day) %>% summarize(!!size_var := mean(!!sym(size_var))), 
                size = 1.25) + 
      labs(x = "Day of measurement", y = y_axis_name) + 
      # limit the viewing to the first 18 days, where we have more measurements per group
      coord_cartesian(xlim = c(0, 18)) + 
      theme(text = element_text(size = 16)) 
  }
# Equivalent to the first plot
tumor_spaghetti("Size", "Tumor Size")
# Equivalent to the second plot
tumor_spaghetti("relative_Size", "Relative Tumor Size")
```



Try to come up with some ways to make the plot fancier. Some suggestions:

(i) Make more informative axis labels using the `labs` or the `scale_x_continuous`/
`scale_y_continuous` functions. 
(ii) Increase the font size using the `theme` function
(iii) Try to layer on to your plot some group-level summary statistics. You 
may be able to do so with a call to `geom_point`.

## Add references

You can create bibliographies in your markdown document. Because I use LaTeX
frequently, I use .bib files to construct mine. I've created one for you and 
dropped it directly in the project folder. Then you can write in-line 
citations as follows:

> The data come from @varna2014tumor

Now knit your document to view your final product!


